MINST Kaggle Digit Recognizer: Contrasting the Random Forest and MLP Neural Network

### 2019-04-15 paper/paper.tex
minor edits to abstract and other things

\begin{abstract}
  In order to tackle the classic MINST dataset, our team implemented a random forest (RF) and a multi-layer perception (MLP) classifier using the \verb+sklearn+ Python library. As expected from a high-dimensional problem such as digit recognition, our RF preformed sub-optimally. Contrasting the accuracy between our training data and testing data suggest severe overfitting. Our final Kaggle submission reached a correct prediction rate of 96.66\%. In order to optimize our MLP NN, we explored the use of stochastic and Adam gradient descent, however found the later consistently outperformed the former. In order to balance accuracy vs. complexity, our cross validation method suggests the use of 1 hidden layer and 512 nodes. Our final Kaggle submission resulted in a correct prediction rate of 97.90\%, suggesting our NN is well-trained. While our MLP preformed considerably better than our RF, the training took approximately 60 times as long, implying the RF may be a desirable option when speed is of utmost importance. 
\end{abstract}

@@

\section{Random Forest}
A Random Forest (RF) is a common starting algorithm for those new to machine learning. This is in part due to it's easy of use, but also due to it's flexibility. While in this assignment it is used for classification, it can be also be used in regression problems [2]. Each tree is formed from a bootstrap sample (random sampling with replacement) then grown very similarly to a decision tree. Rather than searching for the very best split feature, it chooses the best of a random subset.\verb+Sklearn+'s \verb+RandomForestClassifier+ combines the probabilistic prediction of each tree, as this has been shown to preform better than the traditional approach of picking a classification based on a majority vote [4,5].

@@

\subsection{Implementation}
The 784 length vector of input data is normalized to range from 0 to 1 using \verb+sklearn+'s \verb+preprocessing.minmax_scale+. In order to improve our model's predictive power, we contrasted the results from our hyper-parameters \verb+n_estimators+ (number of trees) and  \verb+min_samples_split+ (minimum number of leafs required to split a node). Specifically we allowed 10, 50, 100, and 300 trees to be developed and 2, 4, 8, and 16 as the minimum number of leafs. The later can be thought of as a measure of complexity, with lower minimum resulting in higher complexity. Once the code was prepared, training took approximately 45 seconds. 
\begin{figure}[h]
    \centering
    \begin{minipage}[t]{0.45\textwidth}
        \centering
        \textbf{(A)}
        \includegraphics[width=\textwidth]{CVscoreVsLeafSize.png}
    \end{minipage}
    \begin{minipage}[t]{0.45\textwidth}
        \centering
        \textbf{(B)}
        \includegraphics[width=\textwidth]{CVtestVsLeafSize.png}
    \end{minipage}
\caption{(A,B) Cross Validation of Hyper-parameters for (A) training data and (B) testing data by contrasting mean cross validation score and $\log_2(minleafsize)$ giving the minimum leaf threshold as a measure of complexity. The scores for varying numbers of trees are shown.}
\end{figure}

\subsection{Cross-Validation and Results}
Training and cross-validation was done using \verb+model_selection.GridSearchCV+. In order to visualize the results, Figure 2 contrasts the mean CVS and prediction rates of the training and test datasets. The $\log_2(minleafsize)$ which inversely measures the complexity suggests the more complex models preform outstandingly against the training data but have little impact on the testing data. Adding trees has a similar positive affect on both datasets, however, also decreases exponentially as more trees are added. These results suggest that our model suffers from extreme levels of overfitting. As expected, adding trees does improve our models predictive power but isn't the most practical for our dataset. Our Kaggle submission of 300 trees and a minimum leaf number of 2 gave a prediction rate of 96.6\%

@@

\subsection{Implementation}
In addition to contrasting GD algorithms, the number of hidden nodes were varied between 128, 256, and 512; the number of hidden layers were either 1 or 2; the regularization term $\alpha$ was varied at 0.5, 0.1, 0.001, and 0.0001. The other regularization parameter's for Adam GD were kept at the developer's (Kingma and Lei Ba) recommended defaults: $\epsilon = 10^{-8}$, $\beta_1=0.9$, and $\beta_2=0.999$ [3]. While the learning rate was not varied, an adaptive rate was chosen which divides the current learning rate by 5 with a starting value of $0.001$ [4]. One the code was prepared, training took approximately 46 minutes.

@@

\section{Conclusions}
As expected, our MLP was better suited for handwritten digit recognition. It is possible that data processing to reduce dimensional would help the RF preform better, however computer vision problems are still better left in the hands of NN's and similar models. One advantage of the RF is it's training time was approximately $1/60^{th}$ of our MLP's, however adding addition trees would begin to bridge this gap and further reduce it's run time speed.

>>>>

\begin{abstract}
  In order to tackle the classic MINST dataset, our team implemented a random forest (RF) and a multi-layer perception (MLP) classifier using the \verb+sklearn+ Python library. As expected from a high-dimensional problem such as digit recognition, our RF preformed sub-optimally. Contrasting the accuracy between our training data and testing data suggest severe overfitting. Our final Kaggle submission reached a correct prediction rate of 96.66\%. In order to optimize our MLP NN, we explored the use of stochastic and Adam gradient descent, however found the later consistently outperformed the former. In order to balance accuracy vs. complexity, our cross validation method suggests the use of 1 hidden layer and 512 nodes. Our final Kaggle submission resulted in a correct prediction rate of 97.90\%, suggesting our NN is well-trained. While our MLP preformed considerably better than our RF, the training took approximately 60 times as long, implying the RF may be a desirable option when speed in training is of utmost importance. On the other hand, RF classification time is much greater than that of a NN, which is a nontrivial consideration.
\end{abstract}

@@

\section{Random Forest}
A Random Forest (RF) is a common starting algorithm for those new to machine learning. This is in part due to it's easy of use, but also due to it's flexibility. While in this assignment it is used for classification, it can be also be used in regression problems [2]. Each tree is formed from a bootstrap sample (random sampling with replacement) then grown very similarly to a decision tree. Our random forest is a ``highly random'' random forest which splits data with no discretion to promote model volatility.\verb+Sklearn+'s \verb+RandomForestClassifier+ combines the probabilistic prediction of each tree, as this has been shown to preform better than the traditional approach of picking a classification based on a majority vote [4,5].

@@

\subsection{Implementation}
The 784 length vector of input data is normalized to range from 0 to 1 using \verb+sklearn+'s \verb+preprocessing.minmax_scale+. In order to improve our model's predictive power, we contrasted the results from our hyper-parameters \verb+n_estimators+ (number of trees) and  \verb+min_samples_split+ (minimum number of leafs required to split a node). Specifically we allowed 10, 50, 100, and 300 trees to be developed and 2, 4, 8, and 16 as the minimum number of leaves. The later can be thought of as a measure of complexity, with lower minimum resulting in higher complexity. Once the code was prepared, training took approximately 45 seconds. 
\begin{figure}[h]
    \centering
    \begin{minipage}[t]{0.45\textwidth}
        \centering
        \textbf{(A)}
        \includegraphics[width=\textwidth]{CVscoreVsLeafSize.png}
    \end{minipage}
    \begin{minipage}[t]{0.45\textwidth}
        \centering
        \textbf{(B)}
        \includegraphics[width=\textwidth]{CVtestVsLeafSize.png}
    \end{minipage}
\caption{(A,B) Cross Validation of Hyper-parameters for (A) training data and (B) testing data by contrasting mean cross validation score and $\log_2(minleafsize)$ giving the minimum leaf threshold as a measure of complexity. The scores for varying numbers of trees are shown.}
\end{figure}

\subsection{Cross-Validation and Results}
Training and cross-validation was done using \verb+model_selection.GridSearchCV+. In order to visualize the results, Figure 2 contrasts the mean CVS and prediction rates of the training and test datasets. The $\log_2(minleafsize)$ which inversely measures the complexity suggests the more complex models preform outstandingly against the training data but have little impact on the testing data, a symptom of overfitting. Adding trees has a similar positive affect on both datasets, however, also decreases exponentially as more trees are added. These results suggest that our model suffers from extreme levels of overfitting. As expected, adding trees does improve our models predictive power but isn't the most practical for our dataset. Our Kaggle submission of 300 trees and a minimum leaf number of 2 gave a prediction rate of 96.6\%

@@

\subsection{Implementation}
In addition to contrasting GD algorithms, the number of hidden nodes were varied between 128, 256, and 512; the number of hidden layers were either 1 or 2; the regularization term $\alpha$ was varied at 0.5, 0.1, 0.001, and 0.0001. The other regularization parameter's for Adam GD were kept at the developer's (Kingma and Lei Ba) recommended defaults: $\epsilon = 10^{-8}$, $\beta_1=0.9$, and $\beta_2=0.999$ [3]. While the learning rate was not varied, an adaptive rate was chosen which divides the current learning rate by 5 with a starting value of $0.001$ [4]. One the code was prepared, training took approximately 46 minutes per model. Total training time took approximately 20 hours.

@@

\section{Conclusions}
As expected, our MLP was better suited for handwritten digit recognition. It is possible that data processing to reduce dimensionality would help the RF preform better, however computer vision problems are still better left in the hands of NN's and similar models. One advantage of the RF is it's training time was approximately $1/60^{th}$ of our MLP's, however adding addition trees would begin to bridge this gap and further improve it's classification accuracy.
